---
title: "Lab 5: Rejection Sampling"
author: "Rebecca C. Steorts"
date: "September 22, 2021"
output: 
     pdf_document:
      includes: 
          in_header: custom2.tex
---

# Agenda

We can often end up with posterior distributions that we only know up to a normalizing constant. For example, in practice, we may derive $$p(\theta \mid x) \propto p(x\mid \theta) p(\theta)$$ and find that the normalizing constant $p(x)$ is very difficult to evaluate. Such examples occur when we start building non-conjugate models in Bayesian statistics. 

Given such a posterior, how can we appropriate it's density? One way is using rejection sampling. As an example, let's suppose our resulting posterior distribution is  $$f(x) \propto sin^2(\pi x), x \in [0,1].$$

In order to understand how to approximate the density (normalized) of $f$, we will investigate the following tasks:


1. Plot the densities of $f(x)$ and the Unif(0,1) on the same plot. 
 According to the rejection sampling approach sample from $f(x)$ using the Unif(0,1) pdf as an enveloping function.
2.  Plot a histogram of the points that fall in the acceptance region. Do this for a simulation size of $10^2$ and $10^5$ and report your acceptance ratio. Compare the ratios and histograms.
3.  Repeat Tasks 1 - 3 for  Beta(2,2) as an enveloping function. 

```{r task 1-3 density, echo=F, include=T}
x <- seq(0,1,length=1000)
f <- function(x) {sin(pi*x)^2}
fx <- f(x)

# plot fx and unif on the same plot
plot(x, fx, type='l', lwd=3, col='red', ylab = '', ylim=c(0, 1.6))
# par(new = T)
curve(dunif, add=T, type='l', col='blue', lty=2, lwd=3,
     ylab = '', ylim=c(0, 1.6))
# par(new = T)
curve(dbeta(x, 2,2), add=T, type='l', col='green', lty=3, lwd=3, 
     ylab = '', ylim=c(0, 1.6))
legend("bottom", 
       lty = c(1,2,3), 
       col = c("red", "blue", "green"),
       legend = c(expression(sin(pi*x)^2), "Unif", "Beta(2,2)"))
```


```{r task 1-3 histogram, echo=F, include=T}
set.seed(2022)
# rejection sampling steps
# 1. sample x from g
# 2. sample y|x from unif(0, g(x)); use normalizing to envelope f constant if needed
# 3. if y <= f(x), accept x
rej_sampling <- function(nsim, f, envelope, par1, par2) {
  # step 1
  if (envelope=='unif') {
    x <- runif(nsim, min=par1, max=par2)
    z <- x[lapply(dunif(x,0,1), runif, n=nsim, min=0)[[1]] < f(x)]
    } 
  else if (envelope=='beta') {
    x <- rbeta(nsim, shape1=par1, shape2=par2)
    z <- x[lapply(dbeta(x,2,2), runif, n=nsim, min=0)[[1]] < f(x)]
  }
  # step 2
  # step 3
  acceptance <- length(z)/nsim
  print(acceptance)
  return(z)
}

unif_100 <- rej_sampling(100, f, 'unif', 0, 1)
beta_100 <- rej_sampling(100, f, 'beta', 2, 2)
# unif_100000 <- rej_sampling(10e5, f, 'unif', 0, 1)
# beta_100000 <- rej_sampling(10e5, f, 'beta', 2, 2)
# Terrible code uses too much memory and time -- refer to the provided code

```


4.  Provide the four histograms from Tasks 2 and 3 using the Uniform(0,1) and the Beta(2,2) enveloping proposals. Provide the acceptance ratios. Provide commentary. 
5.  \textcolor{red}{(i) Do you recommend the Uniform or the Beta(2,2) as a better enveloping function (or are they about the same)? (ii) If you were to try and find an enveloping function that had a high acceptance ratio, which one would you try and why? For part (ii), either back up your choice using a small paragraph or (ii) empirically illustrate that it's better. (If you wanted to go above and beyond, you could do both, which is highly encouraged and would really prepare you for the exam!) }


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(1)
```


# Task 1

Plot the densities of $f(x)$ and the Unif(0,1) on the same plot. 

Let's first create a sequence of points from 0 to 1, so that we can have a grid of points for plotting both of the proposed functions. 

```{r}
# grid of points
x <- seq(0, 1, 10^-2)
```

```{r}
fx <- function(x) sin(pi * x)^2
plot(fx, xlim = c(0,1), ylim = c(0,1.5), ylab = "f(x)", lwd = 2)
curve(dunif, add = TRUE, col = "blue", lwd = 2)
curve(dbeta(x,2,2), add = TRUE, col = "red", lwd = 2)
legend("bottom", legend = c(expression(paste("sin(",pi,"x)"^"2")),"Unif(0,1)",
"Beta(2,2)"), col = c("black", "blue", "red"), lty = c(1,1,1), bty = "n", cex = 1.1, lwd = 2)
```

Figure 1: Comparision of the target function and the Unif(0,1) and the Beta(2,2) densities on the same plot.

# Tasks 2 -- 4

According to the rejection sampling approach sample from f(x) using the Unif(0,1) pdf as an enveloping function. In order to do this, we write a general rejection sampling function that also allows us to plot the historams for any simulation size. Finally, our function also allows us to look at task 4 quite easily. 

```{r}
sim_fun <- function(f, envelope = "unif", par1 = 0, par2 = 1, n = 10^2, plot = TRUE){
  
  r_envelope <- match.fun(paste0("r", envelope))
  d_envelope <- match.fun(paste0("d", envelope))
  proposal <- r_envelope(n, par1, par2)
  density_ratio <- f(proposal) / d_envelope(proposal, par1, par2)
  samples <- proposal[runif(n) < density_ratio]
  acceptance_ratio <- length(samples) / n
  if (plot) {
    hist(samples, probability = TRUE, 
         main = paste0("Histogram of ", 
                       n, " samples from ", 
                       envelope, "(", par1, ",", par2,
                       ").\n Acceptance ratio: ",
                       round(acceptance_ratio,2)), 
                       cex.main = 0.75)
  }
  list(x = samples, acceptance_ratio = acceptance_ratio)
}
```

```{r}
par(mfrow = c(2,2), mar = rep(4, 4))
unif_1 <- sim_fun(fx, envelope = "unif", par1 = 0, par2 = 1, n = 10^2) 
unif_2 <- sim_fun(fx, envelope = "unif", par1 = 0, par2 = 1, n = 10^5)
beta_1 <- sim_fun(fx, envelope = "beta", par1 = 2, par2 = 2, n = 10^2) 
beta_2 <- sim_fun(fx, envelope = "beta", par1 = 2, par2 = 2, n = 10^5)
```

Figure 2: Comparision of the output of the rejection sampling for 100 versus 100,000 simulations with uniform and beta distributions as envelope functions.

```{r}
par(mfrow = c(1,1))
```

# Remaining tasks

Be sure to finish any remaining tasks for homework. 

